The fraction of polar aligned circumbinary disks

### 2025-04-23 src/tex/ms.tex
Talk about timescale effects on fpolar in more places

    Circumbinary gas disks that are misaligned to the binary orbital plane evolve toward either a coplanar or a polar-aligned configuration with respect to the binary host. The preferred alignment depends on the dynamics of the disk: whether it undergoes libration or circulating nodal precession, with librating disks evolving to polar inclinations and circulating disks evolving to coplanar.
    We quantify the fraction of binary star systems whose disks are expected to have polar orbits $f_\text{polar}$, extending previous work to include disks with non-zero mass.
    Our results suggest that, for low mass disks, the polar fraction is highly sensitive to the distribution of binary eccentricity with a higher fraction expected for higher binary eccentricities, $f_{\rm polar}\sim e_{\rm b}$. However, for massive discs, the fraction is independent of the binary eccentricity and $f_{\rm polar}\approx 0.37$. The value of $f_\text{polar}$ is always reduced in a population with a greater preference for low initial mutual inclination.
    We also explore the consequences of the finite lifetime and non-zero radial extent of a real disk, which are incomplete evolution and a limit on the relative angular momentum imposed by disk breaking and gravitational instability, respectively. Our findings can be used to make predictions given populations with well-understood distributions of binary eccentricity, initial mutual inclination, and disk angular momentum.
\end{abstract}

@@

Given \RR{$\tau_\text{p}$, the precession timescale in code units, we can calculate the precession timescale in physical units}
\begin{equation}
    \label{eq:tp-j}
    t_\text{p} = \frac{4}{3}\Omega_\text{b}^{-1} \left(\frac{r}{a_\text{b}}\right)^{7/2} \frac{\tau_\text{p}}{f(1-f)\Gamma_2}\, .
\end{equation}
Compared to the precession timescale for a test particle, that is
\begin{equation}
    \label{eq:tp-rat}
    \frac{t_{\text{p,}j}}{t_{\text{p,}j=0}} = \frac{\tau_\text{p}}{4\,\Gamma_1\,\Gamma_2} \, .
\end{equation}

@@

Consider now a wave-like disk with inner radius $r_\text{in}$ and outer radius $r_\text{out}$. If $r_\text{out}$ is sufficiently large, then a break will occur at a radius $r_\text{break}$ that depends on the specific parameters of the disk \citep[see][eq. (9)]{rabago2024}. The two regions can be treated independently of each other. For the inner disk, the outer edge is sufficiently connected to the inner edge, and the whole region precesses together on a characteristic timescale $t_\text{p}(r_\text{in}) < t_\text{p} \lesssim t_\text{p}(r_\text{break})$. The outer disk may break again if $r_\text{out}$ is very large, but either way it will precess on its own characteristic timescale. We now have two independent rings, each with its own precession timescale. Due to the strong dependence of $t_\text{p}$ on $r$, the inner disk is much more likely to reach alignment in a disk lifetime, and the different alignment timescales means the disks will be misaligned from each other, potentially causing a shadow to be cast on the outer disk \RR{\citep[see][]{bohn2022,benisty2023,su2024,zhang2024,zhu2024}}.

@@

We also discuss the limitations of our methods. First, we look at the effect of finite disk lifetime on the ability for a disk to complete its evolution to a stationary state. Disks with long evolutionary timescales remain with orbital parameters near their initial conditions for their entire lifetimes. We also look at limitations imposed by the fact that real disks are not geometrically narrow rings, but have some radial extent. Real disks are likely to break if their radial extent is too large, and will become gravitationally unstable if all of their mass in concentrated in too narrow of a ring. These two effects impose constraints on the relationship between $r_\text{in}$, $r_\text{out}$, and $j$.

>>>>

    Circumbinary gas disks that are misaligned to the binary orbital plane evolve toward either a coplanar or a polar-aligned configuration with respect to the binary host. The preferred alignment depends on the dynamics of the disk: whether it undergoes libration or circulating nodal precession, with librating disks evolving to polar inclinations and circulating disks evolving to coplanar.
    We quantify the fraction of binary star systems whose disks are expected to have polar orbits $f_\text{polar}$, extending previous work to include disks with non-zero mass.
    Our results suggest that, for low mass disks, the polar fraction is highly sensitive to the distribution of binary eccentricity with a higher fraction expected for higher binary eccentricities, $f_{\rm polar}\sim e_{\rm b}$. However, for massive discs, the fraction is independent of the binary eccentricity and $f_{\rm polar}\approx 0.37$. The value of $f_\text{polar}$ is always reduced in a population with a greater preference for low initial mutual inclination.
    We also explore the consequences of the finite lifetime and non-zero radial extent of a real disk, \RR{both of which affect a disk's ability to complete its evolution to a stationary configuration.} Our findings can be used to make predictions given populations with well-understood distributions of binary eccentricity, initial mutual inclination, and disk angular momentum.
\end{abstract}

@@

Given \RR{$\tau_\text{p}$, the precession timescale in units of $\tau$, we can calculate the precession timescale in physical units}
\begin{equation}
    \label{eq:tp-j}
    t_\text{p} = \frac{4}{3}\Omega_\text{b}^{-1} \left(\frac{r}{a_\text{b}}\right)^{7/2} \frac{\tau_\text{p}}{f(1-f)\Gamma_2}\, .
\end{equation}
Compared to the precession timescale for a test particle, that is
\begin{equation}
    \label{eq:tp-rat}
    \frac{t_{\text{p,}j}}{t_{\text{p,}j=0}} = \frac{\tau_\text{p}}{4\,\Gamma_1\,\Gamma_2} \, .
\end{equation}

@@

Consider now a wave-like disk with inner radius $r_\text{in}$ and outer radius $r_\text{out}$. If $r_\text{out}$ is sufficiently large, then a break will occur at a radius $r_\text{break}$ that depends on the specific parameters of the disk \citep[see][eq. (9)]{rabago2024}. The two regions can be treated independently of each other. For the inner disk, the outer edge is sufficiently connected to the inner edge, and the whole region precesses together on a characteristic timescale $t_\text{p}(r_\text{in}) < t_\text{p} \lesssim t_\text{p}(r_\text{break})$. The outer disk may break again if $r_\text{out}$ is very large, but either way it will precess on its own characteristic timescale. We now have two independent rings, each with its own precession timescale. Due to the strong dependence of $t_\text{p}$ on $r$, the inner disk is much more likely to reach alignment in a disk lifetime, and the different alignment timescales means the disks will be misaligned from each other, potentially causing a shadow to be cast on the outer disk \RR{\citep[see][]{bohn2022,benisty2023,su2024,zhang2024,zhu2024}}. \RR{If we choose to measure the polar fraction using the inner ring, then disk breaks will have little effect on $f_\text{polar}$. In practice, however, observed values of $f_\text{polar}$ may depend on whether the technique used to probe the disk configuration is more sensitive to the inner or outer disk.}

@@

We also discuss the limitations of our methods. First, we look at the effect of finite disk lifetime on the ability for a disk to complete its evolution to a stationary state. Disks with long evolutionary timescales remain with orbital parameters near their initial conditions for their entire lifetimes. We also look at limitations imposed by the fact that real disks are not geometrically narrow rings, but have some radial extent. Real disks are likely to break if their radial extent is too large, and will become gravitationally unstable if all of their mass in concentrated in too narrow of a ring. These two effects \RR{shorten the evolutionary timescale and make it more likely that a binary system will host a disk which has reached its stationary configuration, but do not necessarily affect $f_\text{polar}$ as defined in this study}.
